---
title: "modeling_path_efficiency"
output: pdf_document
date: "2025-04-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required Packages

```{r}
library(dplyr)
library(ggplot2)
library(lme4)
library(lmerTest)  # for p-values
library(lattice)
library(emmeans)

```

# Load the data

```{r}
df <- read.csv("~/coding_projects/mouse_data_analysis/data/combined_mouse_data.csv")
head(df)
```

# Make a plot of path efficiency accross trials

```{r}
# Step 1: Average per mouse across experiment sessions
mouse_summary <- df %>%
  group_by(Id, Trial, Genotype, Drug) %>%
  summarize(Mean_Outcome = mean(Outcome, na.rm = TRUE), .groups = "drop")

# Step 2: Compute group mean and standard error
plot_data <- mouse_summary %>%
  group_by(Trial, Genotype, Drug) %>%
  summarize(
    Mean_Path_Efficiency = mean(Mean_Outcome),
    SE = sd(Mean_Outcome) / sqrt(n()),  # Standard Error
    .groups = "drop"
  )

# Step 3: Plotting
ggplot(plot_data, aes(
  x = Trial,
  y = Mean_Path_Efficiency,
  color = Drug,
  group = interaction(Drug, Genotype),
  linetype = Genotype
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) + 
  geom_errorbar(
    aes(ymin = Mean_Path_Efficiency - SE, ymax = Mean_Path_Efficiency + SE),
    width = 0.05
  ) +
  labs(
    title = "Learning Curve: Path Efficiency Over Trials (Combined Dataset)",
    x = "Trial Number",
    y = "Mean Path Efficiency",
    caption = "Error bars represent ±1 standard error of the mean (SEM) across mice."
  ) +
  scale_linetype_manual(values = c("WT" = "dashed", "KI" = "solid")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0.5)  # ← centers the caption
  )

  

```


Based upon the graph, all the mice seem to be learning as their path efficiency is greater in trial 4 than in trial 1. WT also seems to have the greatest rate of improvement, especially when treated with SKF. In contrast, KI mice improve more gradually, and their learning appears less sensitive to drug treatment.

# Convert data types to factors

```{r}
df$Id <- factor(df$Id)
df$Experiment <- factor(df$Experiment)
```

# Fit the LMM
(baseline: Drug: Saline, Genotype: WT, Sex: F, Trial: 1 )
```{r}
lmm <- lmer(
  log(Outcome) ~ Drug*Genotype*Trial + Sex*Trial + (1 | Id/Experiment),
  data = df,
  REML = FALSE
)
summary(lmm)
```

Log transform as ratio is not normal 

# Post Hoc Analysis 



```{r}
# Get estimated slopes of Trial within each Drug × Genotype group
em_trends <- emtrends(lmm, ~ Drug * Genotype, var = "Trial")
summary(em_trends)
```


```{r}
# Contrast the learning slopes between SKF KI and SKF WT
contrast(em_trends, method = list("SKF KI - SKF WT" = c(0, 1, 0, -1)))

```
Clear significance, good! 



Now let us check sex difference. 



```{r}
# Get EMMs for Sex
em_sex <- emmeans(lmm, ~ Sex)

# Contrast the levels of Sex
contrast(em_sex, method = "pairwise")
```
No significant difference 

# Model Diagnostics

We check the plot diagnostics from [here](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf)






## Fitted values vs residuals

```{r}
plot(lmm)
```
Downward trend may indicate slight nonlinearity observed

## scale-location plots
```{r}
plot(lmm, sqrt(abs(resid(.))) ~ fitted(.), type = c("p", "smooth"))
```
Doesn't seem to be an issue with heteroscedasticity 


## Quantile-Quantile plot
```{r}
qqmath(lmm)
```
Some minor deviation from the normality assumption of the residuals. 





