---
title: "modeling_path_efficiency"
output: pdf_document
date: "2025-04-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required Packages

```{r}
library(dplyr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(lattice)
library(emmeans)
library(tidyr)
library(boot)

```

# Load the data

```{r}
df <- read.csv("~/coding_projects/mouse_data_analysis/data/combined_mouse_data.csv")
head(df)
```

# Make a plot of path efficiency accross trials


```{r}
ggplot(plot_data, aes(
  x = Trial,
  y = Mean_Path_Efficiency,
  color = Genotype,
  linetype = Drug,
  group = interaction(Drug, Genotype)
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) + 
  geom_errorbar(
    aes(ymin = Mean_Path_Efficiency - SE, ymax = Mean_Path_Efficiency + SE),
    width = 0.05
  ) +
  labs(
    title = "Learning Curve: Path Efficiency Over Trials (Combined Dataset)",
    x = "Trial Number",
    y = "Mean Path Efficiency",
    caption = "Error bars represent ±1 standard error of the mean (SEM) across mice."
  ) +
  scale_linetype_manual(values = c("Saline" = "dashed", "SKF" = "solid")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0.5)
  )
```





Based upon the graph, all the mice seem to be learning as their path efficiency is greater in trial 4 than in trial 1. WT also seems to have the greatest rate of improvement, especially when treated with SKF. In contrast, KI mice improve more gradually, and their learning appears less sensitive to drug treatment.

# Convert data types to factors

```{r}
df$Id <- factor(df$Id)
df$Experiment <- factor(df$Experiment)
df$Sex <- factor(df$Sex)
df$Drug <- factor(df$Drug)
df$Genotype <- factor(df$Genotype)
df$Trial <- as.numeric(df$Trial)
```





# Fit the LMM
(baseline: Drug: Saline, Genotype: WT, Sex: F, Trial: 1 )
```{r}
lmm <- lmer(
  logit(Outcome) ~ Drug*Genotype*Trial + Sex*Trial + (1 | Id/Experiment),
  data = df,
  REML = FALSE
)
summary(lmm)
```


# Model Based Inference 



```{r}
# Get estimated slopes of Trial within each Drug × Genotype group
em_trends <- emtrends(lmm, ~ Drug * Genotype, var = "Trial")
summary(em_trends)
```


```{r}
# Contrast the learning slopes between SKF KI and SKF WT
contrast(em_trends, method = list("(SKF WT - Saline KI) - (SKF KI - Saline WT)" = c(1, -1, -1, 1)))
```
Clear significance, good! 


## sex differences 
```{r}
# Get EMMs for Sex
em_sex <- emtrends(lmm, ~ Sex, var = "Trial")

# Contrast the levels of Sex
contrast(em_sex, method = "pairwise")
```
No significant difference 

# Model Diagnostics

We check the plot diagnostics from [here](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf)

## Fitted values vs residuals

```{r}
plot(lmm)
```
Downward trend may indicate slight nonlinearity observed

## scale-location plots
```{r}
plot(lmm, sqrt(abs(resid(.))) ~ fitted(.), type = c("p", "smooth"))
```
Doesn't seem to be an issue with heteroscedasticity 


## Quantile-Quantile plot
```{r}
qqmath(lmm)
```
Some minor deviation from the normality assumption of the residuals. 





