---
title: "y_maze"
output:
  pdf_document: default
  html_document: default
date: "2025-05-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load dplyr for filtering
library(dplyr)
library(ggplot2)
library(tidyr)

```

```{r}
df <- read.csv("~/Downloads/Y-Maze.csv")
```

```{r}
head(df)
```




```{r}
# Summarize mean time for each combination
summary_df <- df %>%
  group_by(Genotype, Drug, Novel_Arm) %>%
  summarise(Mean_Time = mean(Time), .groups = "drop")

# Plot
ggplot(summary_df, aes(x = Genotype, y = Mean_Time, fill = as.factor(Novel_Arm))) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~ Drug) +
  labs(
    title = "Mean Time in Novel vs Other Arm",
    x = "Genotype",
    y = "Mean Time (s)",
    fill = "Novel Arm"
  ) +
  scale_fill_manual(values = c("0" = "gray70", "1" = "steelblue"), 
                    labels = c("Other", "Novel")) +
  theme_minimal(base_size = 14)

```



```{r}
library(dplyr)
library(ggplot2)

# Summary stats: group mean + SE
summary_df <- df %>%
  group_by(Genotype, Drug, Novel_Arm) %>%
  summarise(
    Mean_Time = mean(Time, na.rm = TRUE),
    SE = sd(Time, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Plot with bars, black dots for individuals, and error bars
ggplot(summary_df, aes(x = Genotype, y = Mean_Time, fill = as.factor(Novel_Arm))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +

  # Error bars (standard error)
  geom_errorbar(aes(ymin = Mean_Time - SE, ymax = Mean_Time + SE),
                position = position_dodge(width = 0.9), width = 0.2) +

  # Black dots for individual mice
  geom_point(data = df, 
             aes(x = Genotype, y = Time, fill = as.factor(Novel_Arm)),
             color = "black", 
             position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.9),
             size = 1.5, alpha = 0.8, inherit.aes = FALSE) +

  facet_wrap(~ Drug) +
  labs(
    title = "Mean Time in Novel vs Other Arm",
    x = "Genotype",
    y = "Mean Time (s)",
    fill = "Novel Arm"
  ) +
  scale_fill_manual(
    values = c("0" = "gray70", "1" = "steelblue"),
    labels = c("Other", "Novel")
  ) +
  theme_minimal(base_size = 14)

```




```{r}
# Create a new variable for the custom order
df <- df %>%
  mutate(Group = factor(paste(Genotype, Drug),
                        levels = c("WT Saline", "WT SKF", "KI Saline", "KI SKF")))

# Update summary_df accordingly
summary_df <- df %>%
  group_by(Group, Novel_Arm) %>%
  summarise(
    Mean_Time = mean(Time, na.rm = TRUE),
    SE = sd(Time, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Plot with new ordering
ggplot(summary_df, aes(x = Group, y = Mean_Time, fill = as.factor(Novel_Arm))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  geom_errorbar(aes(ymin = Mean_Time - SE, ymax = Mean_Time + SE),
                position = position_dodge(width = 0.9), width = 0.2) +
  geom_point(data = df, 
             aes(x = Group, y = Time, fill = as.factor(Novel_Arm)),
             color = "black", 
             position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.9),
             size = 1.5, alpha = 0.8, inherit.aes = FALSE) +
  labs(
    title = "Mean Time in Novel vs Other Arm",
    x = "Genotype + Drug",
    y = "Mean Time (s)",
    fill = "Novel Arm"
  ) +
  scale_fill_manual(
    values = c("0" = "gray70", "1" = "steelblue"),
    labels = c("Other", "Novel")
  ) +
  theme_minimal(base_size = 14) + theme(axis.line = element_line(color = "black", size = 0.8))

```



```{r}
# Create a new variable for the custom order
df <- df %>%
  mutate(Group = factor(paste(Genotype, Drug),
                        levels = c("WT Saline", "WT SKF", "KI Saline", "KI SKF")))

# Update summary_df accordingly
summary_df <- df %>%
  group_by(Group, Novel_Arm) %>%
  summarise(
    Mean_Time = mean(Time, na.rm = TRUE),
    SE = sd(Time, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Plot with new ordering
ggplot(summary_df, aes(x = Group, y = Mean_Time, fill = as.factor(Novel_Arm))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  geom_errorbar(aes(ymin = Mean_Time - SE, ymax = Mean_Time + SE),
                position = position_dodge(width = 0.9), width = 0.2) +
  geom_point(data = df, 
             aes(x = Group, y = Time, fill = as.factor(Novel_Arm)),
             color = "black", 
             position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.9),
             size = 1.5, alpha = 0.8, inherit.aes = FALSE) +
  labs(
    title = "Mean Time in Novel vs Other Arm",
    x = "Genotype + Drug",
    y = "Mean Time (s)",
    fill = "Novel Arm"
  ) +
  scale_fill_manual(
    values = c("0" = "gray70", "1" = "steelblue"),
    labels = c("Other", "Novel")
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),           # Remove all grid lines
    axis.line = element_line(color = "black", size = 0.8),  # Add black axis lines
    axis.ticks = element_line(color = "black")              # Optional: add axis ticks
  )

```



```{r}
# Ensure Group column is defined
df <- df %>%
  mutate(Group = factor(paste(Genotype, Drug),
                        levels = c("WT Saline", "WT SKF", "KI Saline", "KI SKF")))

# Summary for bar heights
summary_df <- df %>%
  group_by(Group, Novel_Arm) %>%
  summarise(
    Mean_Time = mean(Time, na.rm = TRUE),
    SE = sd(Time, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Compute x positions for dodge manually
df <- df %>%
  mutate(
    Novel_Arm = factor(Novel_Arm),
    x_dodge = as.numeric(Group) + ifelse(Novel_Arm == "0", -0.225, 0.225)
  )

# Now plot with connecting lines per Mouse_ID
ggplot(summary_df, aes(x = Group, y = Mean_Time, fill = Novel_Arm)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  geom_errorbar(aes(ymin = Mean_Time - SE, ymax = Mean_Time + SE),
                position = position_dodge(width = 0.9), width = 0.2) +

  # Connecting lines
  geom_line(data = df,
          aes(x = x_dodge, y = Time, group = interaction(Mouse_ID, Group)),
          color = "black", size = 0.5, alpha = 0.5,
          inherit.aes = FALSE) +

  # Points
  geom_point(data = df,
             aes(x = x_dodge, y = Time, fill = Novel_Arm),
             shape = 21, color = "black", size = 1.5, alpha = 0.8,
             inherit.aes = FALSE) +

  labs(
    title = "Mean Time in Novel vs Other Arm",
    x = "Genotype + Drug",
    y = "Mean Time (s)",
    fill = "Novel Arm"
  ) +
  scale_fill_manual(
    values = c("0" = "gray70", "1" = "steelblue"),
    labels = c("Other", "Novel")
  ) +
  scale_x_discrete(drop = FALSE) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", size = 0.8),
    axis.ticks = element_line(color = "black")
  )

```


```{r}
# 1) Prep your data with numeric x‐positions
dodge_amt <- 0.2
df2 <- df %>%
  mutate(
    Group    = factor(paste(Genotype, Drug),
                      levels = c("WT Saline","WT SKF","KI Saline","KI SKF")),
    Arm      = factor(Novel_Arm, labels = c("Other","Novel")),
    x_center = as.numeric(Group),
    x        = x_center + ifelse(Arm=="Other", -dodge_amt, dodge_amt)
  )

# 2) Subset those mice that have both Other & Novel in each group
line_df <- df2 %>%
  group_by(Group, Mouse_ID) %>%
  filter(n()==2) %>%
  ungroup() %>%
  mutate(Subject_Group = interaction(Group, Mouse_ID))

# 3) Plot violins + paired lines + points
ggplot() +
  # half‐violins on the left/right of each group center
  geom_violin(data = df2,
              aes(x = x, y = Time, group = interaction(Group, Arm), fill = Arm),
              width = dodge_amt*2,
              trim = FALSE,
              alpha = 0.7) +

  # connecting lines
  geom_line(data = line_df,
            aes(x = x, y = Time, group = Subject_Group),
            color = "black", size = 0.4, alpha = 0.6) +

  # individual points
  geom_point(data = df2,
             aes(x = x, y = Time, fill = Arm),
             shape = 21, color = "black", size = 2, alpha = 0.9) +

  # put the group labels back on the x‐axis
  scale_x_continuous(
    breaks = 1:4,
    labels = levels(df2$Group)
  ) +
  scale_fill_manual(values = c(Other = "gray70", Novel = "steelblue")) +

  labs(
    title = "Mean Time in Novel vs Other Arm",
    x     = "Genotype + Drug",
    y     = "Time (s)",
    fill  = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid   = element_blank(),
    axis.line    = element_line(color = "black", size = 0.8),
    axis.ticks   = element_line(color = "black")
  )

```



# Discrimination Index 

```{r}
# Reshape data to wide format to calculate DI per mouse
di_df <- df %>%
  pivot_wider(names_from = Novel_Arm, values_from = Time, names_prefix = "Arm_") %>%
  mutate(
    DI = Arm_1 / (Arm_0 + Arm_1)
  )

```




```{r}
summary_di <- di_df %>%
  group_by(Genotype, Drug) %>%
  summarise(
    Mean_DI = mean(DI, na.rm = TRUE),
    SE = sd(DI, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(Group = factor(paste(Genotype, Drug),
                        levels = c("WT Saline", "WT SKF", "KI Saline", "KI SKF")))

```

```{r}
ggplot(summary_di, aes(x = Group, y = Mean_DI)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) +
  geom_errorbar(aes(ymin = Mean_DI - SE, ymax = Mean_DI + SE), width = 0.2) +
  geom_point(data = di_df %>%
               mutate(Group = factor(paste(Genotype, Drug),
                                     levels = c("WT Saline", "WT SKF", "KI Saline", "KI SKF"))),
             aes(x = Group, y = DI),
             position = position_jitter(width = 0.1), size = 2, alpha = 0.8) +
  labs(
    title = "Discrimination Index by Genotype and Drug",
    x = "Genotype + Drug",
    y = "Discrimination Index (DI = Novel / (Novel + Other))"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", size = 0.8),
    axis.ticks = element_line(color = "black")
  )

```




```{r}
ggplot(di_df %>%
         mutate(Group = factor(paste(Genotype, Drug),
                               levels = c("WT Saline", "WT SKF", "KI Saline", "KI SKF"))),
       aes(x = Group, y = DI)) +
  geom_violin(fill = "steelblue", alpha = 0.7, width = 0.9, trim = FALSE) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 3, fill = "white") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  labs(
    title = "Discrimination Index by Genotype and Drug",
    x = "Genotype + Drug",
    y = "Discrimination Index (DI = Novel / (Novel + Other))"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", size = 0.8),
    axis.ticks = element_line(color = "black")
  )

```




```{r}
di_df <- di_df %>%
  mutate(Group = factor(paste(Genotype, Drug),
                        levels = c("WT Saline", "WT SKF", "KI Saline", "KI SKF")))

ggplot(di_df, aes(x = Group, y = DI, fill = Group)) +
  geom_violin(alpha = 0.7, width = 0.9, trim = FALSE) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.8, aes(color = Group)) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 3, fill = "white") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  labs(
    title = "Discrimination Index by Genotype and Drug",
    x = "Genotype + Drug",
    y = "Discrimination Index (DI = Novel / (Novel + Other))"
  ) +
  scale_fill_manual(values = c(
    "WT Saline"  = "#ADD8E6",  # light blue
    "WT SKF"     = "#4682B4",  # dark blue
    "KI Saline"  = "#F08080",  # light red
    "KI SKF"     = "#8B0000"   # dark red
  )) +
  scale_color_manual(values = c(
    "WT Saline"  = "black",
    "WT SKF"     = "black",
    "KI Saline"  = "black",
    "KI SKF"     = "black"
  )) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", size = 0.8),
    axis.ticks = element_line(color = "black"),
    legend.position = "none"  # optional: hide legend if not needed
  )

```




```{r}
# Filter data for WT only
wt_summary <- summary_di %>% filter(grepl("^WT", Group))
wt_data <- di_df %>% filter(Genotype == "WT") %>%
  mutate(Group = factor(paste(Genotype, Drug),
                        levels = c("WT Saline", "WT SKF")))

# Plot
ggplot(wt_summary, aes(x = Group, y = Mean_DI)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) +
  geom_errorbar(aes(ymin = Mean_DI - SE, ymax = Mean_DI + SE), width = 0.2) +
  geom_point(data = wt_data, aes(x = Group, y = DI),
             position = position_jitter(width = 0.1), size = 2, alpha = 0.8) +
  labs(
    title = "Discrimination Index for WT Mice",
    x = "Drug",
    y = "Discrimination Index (DI = Novel / (Novel + Other))"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", size = 0.8),
    axis.ticks = element_line(color = "black")
  )
```




```{r}
# Filter data for KI only
ki_summary <- summary_di %>% filter(grepl("^KI", Group))
ki_data <- di_df %>% filter(Genotype == "KI") %>%
  mutate(Group = factor(paste(Genotype, Drug),
                        levels = c("KI Saline", "KI SKF")))

# Plot
ggplot(ki_summary, aes(x = Group, y = Mean_DI)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) +
  geom_errorbar(aes(ymin = Mean_DI - SE, ymax = Mean_DI + SE), width = 0.2) +
  geom_point(data = ki_data, aes(x = Group, y = DI),
             position = position_jitter(width = 0.1), size = 2, alpha = 0.8) +
  labs(
    title = "Discrimination Index for KI Mice",
    x = "Drug",
    y = "Discrimination Index (DI = Novel / (Novel + Other))"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", size = 0.8),
    axis.ticks = element_line(color = "black")
  )

```






```{r}
df$Mouse_ID <- factor(df$Mouse_ID)
df$Drug <- factor(df$Drug)
df$Genotype <- factor(df$Genotype)
df$Novel_Arm <- factor(df$Novel_Arm)
df$Group <- interaction(df$Drug, df$Genotype)

# Get unique groups
groups <- unique(df$Group)

# Store results
results <- data.frame(Group = character(), p = numeric(), t = numeric(), df = numeric(),
                      mean_diff = numeric(), shapiro_p = numeric(), stringsAsFactors = FALSE)

# Loop through each group
for (g in groups) {
  cat("\n---", g, "---\n")
  sub_df <- subset(df, Group == g)
  
  # Reshape to wide format
  wide <- reshape(sub_df, idvar = "Mouse_ID", timevar = "Novel_Arm", direction = "wide")
  
  if (all(c("Time.0", "Time.1") %in% names(wide))) {
    diffs <- wide$Time.1 - wide$Time.0
    shapiro_test <- shapiro.test(diffs)
    #test <- t.test(wide$Time.1, wide$Time.0, paired = TRUE)
    test <- t.test(wide$Time.1, wide$Time.0, paired = TRUE, alternative = "greater")

    cat("Shapiro-Wilk p-value:", round(shapiro_test$p.value, 4), "\n")
    print(test)
    
    results <- rbind(results, data.frame(
      Group = as.character(g),
      p = test$p.value,
      t = test$statistic,
      df = test$parameter,
      mean_diff = mean(diffs),
      shapiro_p = shapiro_test$p.value
    ))
  } else {
    cat("Skipping group due to insufficient data\n")
  }
}

# Adjust p-values
results$adjusted_p <- p.adjust(results$p, method = "holm")

# Print final result table
print(results)

```

## Paired t-tests: Novel Arm Preference

We tested whether mice spent significantly more time in the **novel arm** than in the **familiar arm** within each Drug × Genotype group using **one-sided paired t-tests**.

### Hypotheses

Let $\mu_{\text{novel}}$ be the mean time spent in the novel arm, and $\mu_{\text{familiar}}$ the mean time in the familiar arm.

-   **Null hypothesis**: $H_0: \mu_{\text{novel}} \leq \mu_{\text{familiar}}$
-   **Alternative hypothesis**: $H_1: \mu_{\text{novel}} > \mu_{\text{familiar}}$

We use `alternative = "greater"` in `t.test()`.

### Assumptions

-   **Paired measurements**: Each mouse provides both arm 0 and arm 1 observations.
-   **Normality of differences**: Checked using the [Shapiro–Wilk test](https://en.wikipedia.org/wiki/Shapiro%E2%80%93Wilk_test).\
    If p \> 0.05, the normality assumption holds.
-   **Multiple comparisons**: Corrected using the [Holm–Bonferroni method](https://en.wikipedia.org/wiki/Holm%E2%80%93Bonferroni_method).

### Results

Overall, only the **SKF + WT** group spent a significantly greater amount of time in the novel arm (*p* = 0.0066, Holm-adjusted *p* = 0.0266).
